---
title: "Data Analysis and Visualization Project"
author: "Tiffany Pang"
date: "4/28/2017"
output:
  html_document: default
---


## **Creating a new Category column**

***In order to analyze and visualize data in the FNDDS 2011-2012 Foods database, Bob Horton’s starter script is used to create the database on the machine from flat files.***

```{r starter_script, warning=FALSE, message=FALSE}

data_dir <- "FNDDS_2011"

fortification <- c(`0`="none", `1`="fortified_product", `2`="contains fortified ingredients")

fndds_tables <- list(
	AddFoodDesc = list(
			title="Additional Food Descriptions",
			column_types=c(
				food_code="integer", # foreign key
				seq_num="integer", 
				start_date="date", 
				end_date="date", 
				additional_food_description="text"),
			sep="^"
		),
	FNDDSNutVal = list(
			title="FNDDS Nutrient Values",
			column_types=c(
				food_code="integer",
				nutrient_code="integer",	# Nutrient Descriptions table
				start_date="date", 
				end_date="date", 
				nutrient_value="double"
				),
			sep="^"
		),
	FNDDSSRLinks = list(
			title="FNDDS-SR Links",	# see p34 of fndds_2011_2012_doc.pdf
			column_types=c(
				food_code="integer",
				start_date="date", 
				end_date="date", 
				seq_num="integer",
				sr_code="integer",
				sr_descripton="text",
				amount="double",
				measure="char[3]",	# lb, oz, g, mg, cup, Tsp, qt, fluid ounce, etc
				portion_code="integer",
				retention_code="integer",
				flag="integer",
				weight="double",
				change_type_to_sr_code="char[1]",	# D=data change; F=food change
				change_type_to_weight="char[1]",
				change_type_to_retn_code="char[1]"
				),
			sep="^"
		),
	FoodPortionDesc = list(
			title="Food Portion Descriptions",
			column_types=c(
				portion_code="integer", 	# foreign key
				start_date="date",
				end_date="date",
				portion_description="text",
				change_type="char[1]"
			),
			sep="^"
		),
	FoodSubcodeLinks = list(
			title="Food code-subcode links",
			column_types=c(
				food_code="integer",
				subcode="integer",
				start_date="date",
				end_date="date"
				),
			sep="^"
		),
	FoodWeights = list(
			title="Food Weights",
			column_types=c(
				food_code="integer",	# foreign key
				subcode="integer",
				seq_num="integer",
				portion_code="integer",	# food portion description id
				start_date="date",
				end_date="date",
				portion_weight="double",	# missing values = -9
				change_type="char[1]"	# D=data change, F=food change
				),
			sep="^"
		),
	MainFoodDesc = list(
			title="Main Food Descriptions",
			column_types=c(
				food_code="integer", 
				start_date="date", 
				end_date="date", 
				main_food_description="character", 
				fortification_id="integer"),
			sep="^"
		),
	ModDesc = list(
			title="Modifications Descriptons",
			column_types=c(
				modification_code="integer",
				start_date="date", 
				end_date="date", 
				modification_description="text",
				food_code="integer"
				
				),
			sep="^"
		),
	ModNutVal = list(
			title="Modifications Nutrient Values",
			column_types=c(
				modification_code="integer",
				nutrient_code="integer",
				start_date="date", 
				end_date="date", 
				nutrient_value="double"
				),
			sep="^"
		),
	MoistNFatAdjust = list(
			title="Moisture & Fat Adjustments",	# to account for changes during cooking
			column_types=c(
				food_code="integer",
				start_date="date", 
				end_date="date", 
				moisture_change="double",
				fat_change="double",
				type_of_fat="integer"	# SR code or food code				
				),
			sep="^"
		),
	NutDesc = list(
			title="Nutrient Descriptions",
			column_types=c(
				nutrient_code="integer",
				nutrient_description="text",
				tagname="text",
				unit="text",
				decimals="integer"	# decimal places
				),
			sep="^"
		),
	SubcodeDesc = list(
			title="Subcode Descriptions",
			column_types=c(
				subcode="integer",	# key; 0=use default gram weights
				start_date="date",
				end_date="date",
				subcode_description="text"
				),
			sep="^"
		)
)

# flat file to a data frame: call for each table
assign_data_frame <- function(tbl_name){
	tbl <- read.table(
		file.path(data_dir, paste0(tbl_name, ".txt")), 
		sep="^",
		quote="~",
		stringsAsFactors=FALSE)
	# drop last (empty) column
	tbl <- tbl[1:(length(tbl)-1)]
	names(tbl) <- names(fndds_tables[[tbl_name]][["column_types"]])
	assign(tbl_name, tbl, envir = .GlobalEnv)
}

# flat file to database
fndds2sqlite <- function(data_dir, table_details, sqlite_filename){

	library("RSQLite")
	con <- dbConnect(SQLite(), sqlite_filename)

	for (tbl_name in names(table_details)){
		file_name <- paste0(tbl_name, ".txt")
		assign_data_frame(tbl_name)
		tbl <- get(tbl_name)
		dbWriteTable(con, tbl_name, tbl, row.names = FALSE)
	}
	
	dbDisconnect(con)
}

fndds2sqlite("FNDDS_2011", fndds_tables, "fndds.sqlite")
library(DBI)

for (tbl in c("FNDDSNutVal", "MainFoodDesc", "NutDesc"))
	assign_data_frame(tbl)

library(dplyr)
library(tidyr)

# Make a simplified selection of foods.
# TO DO: have MainFoodDesc be a tbl sourced from SQLite.
get_selected_foods <- function(){
	# Pull out all "Not Further Specified" foods as a wide selection of reasonably generic items.
	generics <- MainFoodDesc %>% 
		filter( grepl(", NFS", main_food_description )) %>%
		filter(!grepl("infant formula", main_food_description, ignore.case = TRUE ) )

	# Raw fruits
	# Berries are covered by "Berries, raw, NFS" and "Berries, frozen, NFS"
	fruits <- MainFoodDesc %>% 
		filter( grepl("^6", food_code) ) %>%
		filter( grepl("^([^,\\(]+), raw$", main_food_description) ) %>% 
		filter( !grepl("berries", main_food_description) )

	# Raw vegetables
	# Potatoes are covered by "White potato, NFS", "Sweet potato, NFS", etc.
	vegetables <- MainFoodDesc %>% 
		filter( grepl("^7", food_code) ) %>%
		filter(!grepl("potato", main_food_description)) %>%
		filter( grepl(", raw$", main_food_description))

	# 4="legumes, nuts, and seeds"
	nuts_and_seeds <- MainFoodDesc %>% 
		filter( grepl("^4", food_code) ) %>%
		mutate( firstWord = strsplit(main_food_description, " ")[[1]][1] )
	
	# Selected alcoholic beverages
	# All alcoholic beverages: grepl("^93", food_code))
	# "Cocktail, NFS" already gives us "Cocktail"
	alcoholic_beverages <- MainFoodDesc %>% 
		filter( main_food_description %in% c("Beer", "Wine, table, red", "Wine, table, white", 
			"Whiskey", "Gin", "Rum", "Vodka") )

	# Collect them all into one table
	rbind(generics, fruits, vegetables, alcoholic_beverages) %>%
		select( food_code, main_food_description, fortification_id )  %>% 
		filter( nchar(main_food_description) < 20 ) %>%
		mutate( main_food_description = gsub("(, NFS|, raw)", "", main_food_description) ) 

}

foods <- get_selected_foods()	# 163 items

```

***The following code is then used to create a new column named Category based on Appendix E. Food/Beverage Coding Scheme, and appended to food_nutrient_df.***

```{r append catagory column, warning = FALSE, message=FALSE}
library(sqldf)

long_food_nutrients_food_code <-
  sqldf(
    "SELECT f.food_code, nd.nutrient_description, nv.nutrient_value
    FROM foods f
    INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code
    INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code"
  )

food_code_dataframe <- spread(long_food_nutrients_food_code, food_code, nutrient_value, fill=0)
food_code_mat <- t(as.matrix(food_code_dataframe[-1]))
colnames(food_code_mat) <- food_code_dataframe$nutrient_description

food_code <- as.integer(row.names(food_code_mat))
Category <- rep("dairy", length(food_code))
temp_df <- data.frame(food_code, Category, stringsAsFactors=F)

temp_df$Category[temp_df$food_code>=94000000] <- "protein powder"
temp_df$Category[temp_df$food_code<94000000] <- "alcohol"
temp_df$Category[temp_df$food_code<93000000] <- "sugars"
temp_df$Category[temp_df$food_code<90000000] <- "fats"
temp_df$Category[temp_df$food_code<80000000] <- "vegetables"
temp_df$Category[temp_df$food_code<70000000] <- "fruits"
temp_df$Category[temp_df$food_code<60000000] <- "grains"
temp_df$Category[temp_df$food_code<50000000] <- "legumes,nuts,seeds"
temp_df$Category[temp_df$food_code<40000000] <- "eggs"
temp_df$Category[temp_df$food_code<30000000] <- "meat,fish"
temp_df$Category[temp_df$food_code<20000000] <- "dairy"

food_nutrient_df <- as.data.frame(food_code_mat, stringsAsFactors = FALSE)
food_nutrient_df <- cbind(food_nutrient_df, Category=temp_df$Category)
```

***Necessary changes are then made to the dataframe for further exploratory data analysis***

```{r warning = FALSE, message=FALSE}
# remove the first 19 columns of the dataframe
food_df <- food_nutrient_df[, -c(1:19)]

# replace spaces in column names with underscores
library(stringr)
colnames(food_df) <- str_replace_all(colnames(food_df),"[[:punct:]\\s]+","_")
```

***The following code was used to find the count of each food category.***
```{r category count}
as.data.frame(table(food_df$Category))
```

***Check the class of all variables.***
```{r check class}
as.data.frame(sapply(food_df, class))
```

***Summary of the data.***
```{r data summary}
summary(food_df)
```
\newpage

## **Data Visualization with ggplot**

```{r}
library(ggplot2)

# Visualize the count of each Food Category
g <- ggplot(food_df, aes(x=Category)) 
g + geom_bar(aes(fill = Category)) + 
  labs(title = "Count of Each Food Category\n") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
\newpage
```{r}
# Boxplot of Total Fat by Food Category
ggplot(food_df, aes(x=Category, y=Total_Fat)) + # categorical variable on x-axis
  geom_boxplot(aes(fill = Category)) +
  labs(title = "Total Fat by Food Category\n") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
***The category 'fats' contains the highest total_fat, followed by 'legumes,nuts,seeds'. 'dairy' and 'meat,fish' categories contain the same amount of total_fat.***
\newpage
```{r}
# Boxplot of Carbohydrate by Food Category
ggplot(food_df, aes(x=Category, y=Carbohydrate)) + # categorical variable on x-axis
  geom_boxplot(aes(fill = Category)) + coord_flip() +
  labs(title = "Carbohydrate by Food Category\n") +
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
***Sugars contain the highest amount of carbohydrate, followed by grains.***
\newpage
```{r}
# Interleaved histogram of fiber_total_dietary by Food Category
ggplot(food_df, aes(x=Fiber_total_dietary, fill=Category)) +
  geom_histogram(position = "dodge", binwidth = 4) +
  labs(title = "Fiber_total_dietary by Food Category\n") +
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
***Grains, legumes, nuts, seeds and fruits contain the highest total dietary fiber and alcohol, diary and fats contain no fiber.***
\newpage
```{r}
# Energy as a function of Total_Fat for each Food Category
ggplot(food_df, aes(x = Total_Fat, y = Energy, colour = Category)) +
  geom_point() +
  facet_wrap( ~ Category) + 
  labs(title="Energy as a function of Total_Fat for each Food Category\n", x="Total_Fat", y="Energy") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=12, face="bold", color="darkgreen"))
```
***Fats category has the highest amount of total_fat and energy, whereas vegetables category has low total_fat and also low energy.***
\newpage
```{r}
# Scatterplot of Energy as a Function of Water by Each Food Category
ggplot(food_df, aes(x=Water,y=Energy)) +
  geom_point(aes(fill=Category, color=Category), size=2) +
  labs(title="Energy as a Function of Water by Each Food Category\n") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=12, face="bold", color="darkgreen"))
```
***As the water content of the food goes up, it contains lower energy. Water and energy are inversely correlated. Again, here vegetables have the highest water content and provide lowest energy.***
\newpage
```{r}
# Scatterplot of Cholesterol as a function of Fatty_acids_total_saturated
ggplot(food_df, aes(x=Fatty_acids_total_saturated, y=Cholesterol)) + 
  geom_point(aes(shape=Category, color=Category), size=3.5) +
  scale_shape_manual(values=c(6, 4, 18, 17, 5, 16, 15, 8, 10, 11)) + 
  labs(title="Scatterplot of Cholesterol as a function of Fatty_acids_total_saturated for each Food Category\n") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=12, face="bold", color="darkgreen"))
```
***Fats category has the highest total saturated fatty acids and cholesterol. Meat and fish have second highest cholesterol, but dairy has the second highest total saturated fatty acids.***
\newpage
```{r}
# Sugars_total for each Food Category
ggplot(food_df, aes(x=Sugars_total, y=Category)) + 
  geom_point(aes(color=Category)) + 
  labs(title="Sugars_total for each Food Category\n") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
***Sugars category has the highest total sugars, followed by fruits. Protein powder and fats have no sugar content.***
\newpage
```{r}
# Amount of Calcium content for Each Food Category
ggplot(food_df, aes(x=Category, y=Calcium))+
  geom_bar(stat="identity", aes(fill=Category, color=Category)) +
  coord_polar(theta = "x", direction=1 ) + 
  labs(title="Amount of Calcium content for Each Food Category\n") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=14, face="bold", color="darkgreen"))
```
***Vegetables contains the most calcium, followed by grains and dairy. Alcohol, fats and sugars do not seem to contain any calcium.***
\newpage
```{r warning=FALSE}
# Histogram of Protein by each Food Category
ggplot(food_df, aes(Protein))+ 
  geom_histogram(aes(color=Category, fill=Category)) +
  facet_grid(Category ~ .) + 
  labs(title="Histogram of Protein by each Food Category\n") + 
  theme_bw() +
  theme(plot.title=element_text(hjust=0.5, size=16, face="bold", color="darkgreen"))
```
***Protein powder contains the highest protein amount, followed by meat, fish and legumes, nuts, seeds. Alcohol, fats and sugars do not contain any protein.***
\newpage

## **SQL Queries** 

***Food_code column is added to the dataframe, so it can be used for joining with other tables for SQL queries.***
```{r append Food_code column}
df <- cbind(food_df, food_code=temp_df$food_code)
```

***The following SQL queries are written to understand and explore the dataset.***

```{r}
library(sqldf)
#1
sqldf("SELECT Category, count(*) AS counts FROM df GROUP BY Category")
```

```{r}
#2
# number of food high in Calcium
sqldf("SELECT Category, count(*) as counts FROM df WHERE Calcium > 100
      GROUP BY Category")
```

```{r}
#3
# Percentage of each food category in regards to all food categories 
sqldf(
  "SELECT Category, ROUND((cast(sub_query.counts as float)/sub_query.total)*100, 2) AS
  Percentage_of_Each_Category 
  FROM (SELECT Category, count(*) AS counts, 
  (SELECT count(*) FROM df) AS total FROM df GROUP BY Category) AS sub_query 
  ORDER BY Percentage_of_Each_Category DESC")
```

```{r}
#4
# Averages of the nutrients 
sqldf(
  "SELECT Category, AVG(Fatty_acids_total_monounsaturated) AS Avg_monounsat, 
  Avg(Fatty_acids_total_polyunsaturated) AS Avg_polyunsat, 
  Avg(Fatty_acids_total_saturated) AS Avg_sat
  FROM df
  GROUP BY Category")
```

```{r}
sqldf(
  "SELECT Category, AVG(Calcium), AVG(Magnesium), AVG(Iron), AVG(Phosphorus), 
  AVG(Potassium), AVG(Sodium), AVG(Zinc)
  FROM df
  GROUP BY Category")
```

```{r}
sqldf(
  "SELECT Category, AVG(Energy), AVG(Folate_total), AVG(Sugars_total), 
  AVG(Total_Fat), AVG(Water)
  FROM df
  GROUP BY Category")
```

```{r}
#5
# Max, min and avgerages of some nutrients 
sqldf(
  "SELECT Category, MAX(Cholesterol), MIN(Cholesterol), AVG(Cholesterol)
  FROM df
  GROUP BY Category
  ORDER BY Category")
```

```{r}
sqldf(
  "SELECT Category, MAX(Protein), MIN(Protein), AVG(Protein)
  FROM df
  GROUP BY Category
  ORDER BY MAX(Protein)")
```

```{r}
sqldf(
  "SELECT Category, MAX(Fiber_total_dietary), MIN(Fiber_total_dietary), 
  AVG(Fiber_total_dietary)
  FROM df
  GROUP BY Category
  ORDER BY AVG(Fiber_total_dietary) DESC")
```

```{r}
#6
# Sum of some food nutrients 
sqldf(
  "SELECT Category, SUM(Total_Fat), SUM(Cholesterol), SUM(Carbohydrate), 
  SUM(Protein), SUM(Sugars_total) 
  FROM df 
  GROUP BY Category")
```

```{r}
#7
# Distinct nutrtient descriptions 
sqldf("SELECT DISTINCT(nutrient_description) 
      FROM NutDesc
      LIMIT 10")
```

```{r}
#8
# Main and additional food description
head(
  sqldf(
    "SELECT f.main_food_description, afd.additional_food_description
    FROM foods f
    INNER JOIN AddFoodDesc afd ON f.food_code = afd.food_code")
)
```

```{r}
#9
# Additional food description that contain words with ES
sqldf(
  "SELECT f.main_food_description, afd.additional_food_description
  FROM foods f
    INNER JOIN AddFoodDesc afd ON f.food_code = afd.food_code
    WHERE additional_food_description LIKE '%ES%'
    ORDER BY main_food_description")
```

```{r}
#10
# Food with the highest Cholesterol and Total Fat
head(
  sqldf(
    "SELECT f.main_food_description, nd.nutrient_description, nv.nutrient_value
    FROM foods f
      INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code
      INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code
      WHERE nutrient_description IN ('Cholesterol', 'Total Fat')
      ORDER BY nutrient_value DESC")
)
```

```{r}
#11
# food with the max and avg nutrient_value in any one nutrient description, 
#  ordered by max nutrient_value 
sqldf(
  "SELECT f.main_food_description, d.Category, nd.nutrient_description, 
  AVG(nv.nutrient_value), MAX(nv.nutrient_value)
  FROM FNDDSNutVal nv
    INNER JOIN foods f ON f.food_code = nv.food_code
    INNER JOIN df d ON f.food_code = d.food_code
    INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code
    GROUP BY main_food_description
    ORDER BY MAX(nv.nutrient_value) DESC
    LIMIT 10")
```

```{r}
#12 
# High Energy, low cholesterol and low sugar
sqldf(
  "SELECT f.main_food_description, d.Category, d.Energy, d.Cholesterol, d.Sugars_total
  FROM foods f
    INNER JOIN df d ON f.food_code = d.food_code
    WHERE d.Energy >= 400 AND d.Cholesterol < 200 AND d.Sugars_total < 15
    ORDER BY d.Energy DESC")
```

```{r}
# High carb low fat food
sqldf(
  "SELECT DISTINCT f.main_food_description, d.Category, d.Carbohydrate, d.Total_Fat
  FROM foods f
    INNER JOIN df d ON f.food_code = d.food_code
    WHERE Carbohydrate > 50 AND Total_Fat < 50
    GROUP BY main_food_description
    ORDER BY Carbohydrate DESC")
```

```{r}
#13
# Food highest in Energy 
sqldf("SELECT f.main_food_description, afd.additional_food_description, 
  d.Category, nd.nutrient_description, nv.nutrient_value 
	FROM foods f 
      INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code 
      INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code
      INNER JOIN AddFoodDesc afd ON f.food_code = afd.food_code
      INNER JOIN df d ON f.food_code = d.food_code
      WHERE nutrient_description = 'Energy'
      ORDER BY nutrient_value DESC
      LIMIT 5") 
```

```{r}
# Food highest in Vitamin C
sqldf("SELECT f.main_food_description, afd.additional_food_description, 
  d.Category, nd.nutrient_description, nv.nutrient_value 
	FROM foods f 
      INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code 
      INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code
      INNER JOIN AddFoodDesc afd ON f.food_code = afd.food_code
      INNER JOIN df d ON f.food_code = d.food_code
      WHERE nutrient_description = 'Vitamin C'
      ORDER BY nutrient_value DESC
      LIMIT 5") 
```

```{r}
# Food highest in Sugars, total
sqldf("SELECT f.main_food_description, afd.additional_food_description, 
  d.Category, nd.nutrient_description, nv.nutrient_value 
	FROM foods f 
      INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code 
      INNER JOIN NutDesc nd ON nv.nutrient_code = nd.nutrient_code
      INNER JOIN AddFoodDesc afd ON f.food_code = afd.food_code
      INNER JOIN df d ON f.food_code = d.food_code
      WHERE nutrient_description = 'Sugars, total'
      ORDER BY nutrient_value DESC
      LIMIT 5") 
```

```{r}
#14
# Max monounsat, polyunsat, and sat in 'fats', 'legumes,nuts,seeds', 'grains' catagories, 
#  ordered by max_sat descending
sqldf(
  "SELECT sub_query.main_food_description, sub_query.Category, sub_query.Max_monounsat, 
  sub_query.Max_polyunsat, sub_query.Max_sat
  FROM
    (SELECT f.main_food_description, d.Category, MAX(d.Fatty_acids_total_monounsaturated) 
    AS Max_monounsat, MAX(d.Fatty_acids_total_polyunsaturated) AS Max_polyunsat, 
    MAX(d.Fatty_acids_total_saturated) AS Max_sat
    FROM foods f
    INNER JOIN df d ON f.food_code = d.food_code
    GROUP BY main_food_description) AS sub_query
  WHERE Category IN ('fats', 'legumes,nuts,seeds', 'grains')
  ORDER BY Max_sat DESC
  LIMIT 5")
```

```{r}
#15
# Total nutrient value for each food description
sqldf("SELECT f.main_food_description, d.Category, 
  SUM(nv.nutrient_value) AS total_nutrient_value
	FROM foods f 
      INNER JOIN FNDDSNutVal nv ON f.food_code = nv.food_code 
      INNER JOIN df d ON f.food_code = d.food_code
      GROUP BY main_food_description
      ORDER BY total_nutrient_value DESC
      LIMIT 10")
```

```{r}
#16
# Total nutrient value for each food category
sqldf("SELECT d.Category, SUM(nv.nutrient_value) AS total_nutrient_value
	FROM df d
      INNER JOIN FNDDSNutVal nv ON d.food_code = nv.food_code 
      GROUP BY Category
      ORDER BY total_nutrient_value DESC")
```

***The following SQL queries are used for subsetting different food category datasets.***
```{r}
dairy_df <- sqldf("SELECT * FROM df WHERE Category = 'dairy'")
grains_df <- sqldf("SELECT * FROM df WHERE Category = 'grains'") 
meat_df <- sqldf("SELECT * FROM df WHERE Category = 'meat,fish'") 
nuts_df <- sqldf("SELECT * FROM df WHERE Category = 'legumes,nuts,seeds'") 
fats_df <- sqldf("SELECT * FROM df WHERE Category = 'fats'") 
fruit_veg_df <- sqldf("SELECT * FROM df WHERE Category IN ('fruits', 'vegetables')") 
```


***The SQL queries help us understand the ranges of nutritional content among various food categories, and also which categories/food are high in certain nutritional values. Next, some t-tests are run to infer difference in mean nutrient value for the population of foods from which this sample is drawn.***

## **T-Tests**

```{r}
#1
t.test(dairy_df$Protein, meat_df$Protein)
```

***p-value is greater than 0.05, fail to reject null hypothesis.***
***There is no difference in mean nutrient value for these two categories.***

```{r}
#2
t.test(nuts_df$Total_Fat, grains_df$Total_Fat)
```

***p-value is less than 0.05, reject null hypothesis.***
***There is a difference in mean nutrient value for these two categories.***

```{r}
#3
t.test(nuts_df$Energy, fats_df$Energy)
```

***p-value is greater than 0.05, fail to reject null hypothesis.***
***There is no difference in mean nutrient value for these two categories.***

```{r}
#4
t.test(fruit_veg_df$Fiber_total_dietary, grains_df$Fiber_total_dietary)
```

***p-value is greater than 0.05, fail to reject null hypothesis.***
***There is no difference in mean nutrient value for these two categories.***

#### ***T-tests reveal that some food catogories contain similar mean value of nutritional contents.***

## **Correlations**

```{r}
library(reshape2)
library(scales)
# calculate correlation matrix
correlationMatrix <- cor(df[, -c(47, 48)])
# melt it into the long format
foodMelt <- melt(correlationMatrix, varnames=c("x", "y"), value.name="Correlation")
# order it according to the correlation
foodMelt <- foodMelt[order(foodMelt$Correlation), ]
```

```{r fig.width=16, fig.height=14}
# plot of correlation heatmap
ggplot(foodMelt, aes(x=x, y=y)) + 
  geom_tile(aes(fill = Correlation)) + 
  scale_fill_gradient2(low = muted("deeppink4"), mid = "white", 
                       high = "blue")+
  labs(title="Heatmap of the Correlation of Food Nutrients") + 
  theme_minimal()+ 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
  theme(plot.title=element_text(hjust=0.5, size=20, face="bold", 
                                color="darkgreen")) + 
  xlab("Food Nutrients") + 
  ylab("Food Nutrients")
```

***The following code is then used to find the highly correlated variables (correlation > 0.5).***

```{r}
library(mlbench)
library(caret)
highlyCorrelated <- findCorrelation(correlationMatrix, cutoff=0.5, names = TRUE)
highlyCorrelated
```

## **Logistic Regression**

```{r warning=FALSE}
# remove food_code column 
fruit_veg_df$food_code <- NULL
# set fruits as 0 and vegetables as 1
fruit_veg_df$Category <- ifelse(fruit_veg_df$Category=='fruits', 0, 1)

fit_main <- glm(Category ~ ., family = binomial(), data = fruit_veg_df)
```
***Summary(fit_main):***   
***Null deviance: 1.2191e+02  on 89  degrees of freedom
Residual deviance: 3.3005e-09  on 51  degrees of freedom
AIC: 78***

```{r warning=FALSE}
# Fit a regression model for the null model: Category as a function of the intercept only.
fit_null <- glm(Category ~ 1, family = binomial(), data = fruit_veg_df)
```
***Summary(fit_null):***
***Null deviance: 121.91  on 89  degrees of freedom
Residual deviance: 121.91  on 89  degrees of freedom
AIC: 123.91***  

***Next, step function is used for variable selection. The step function iterates through possible models, and return the optimal model with the lowest AIC.***
```{r echo=FALSE, warning=FALSE, results='hide'}
fit_step = step(fit_null, scope=list(lower=fit_null,
                                     upper=fit_main),direction="both")

```

***The optimal model (AIC = 14) returned from the step function is then fitted as fit_final.***
```{r warning=FALSE}
# final model returned from step function
fit_final <- glm(Category ~ Sugars_total + Sodium + Thiamin + 
                 Fatty_acids_total_monounsaturated + Choline_total + Vitamin_B_6,
                 family = binomial(), data = fruit_veg_df)
summary(fit_final)
```


```{r warning=FALSE}
# Written by Andy Field
logisticPseudoR2s <- function(LogModel) {
  dev <- LogModel$deviance 
  nullDev <- LogModel$null.deviance 
  modelN <-  length(LogModel$fitted.values)
  R.l <-  1 -  dev / nullDev
  R.cs <- 1- exp ( -(nullDev - dev) / modelN)
  R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
  cat("Pseudo R^2 for logistic regression\n")
  cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
  cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
  cat("Nagelkerke R^2           ", round(R.n, 3),    "\n")
}
logisticPseudoR2s(fit_final)

# Anova Test 
a_mcv <- anova(fit_final)
a_mcv
```
